Detection of Oscillations in Aldebaran with Ground-Based Observations

### 2018-01-12 paper/aldebaran.tex
appendix ref

It is easy to observe Aldebaran and similarly-bright stars with ground-based spectroscopic instruments, typically only requiring short exposures that can be obtained even under adverse observing conditions. There is indeed a considerable archive of such observations already, as a legacy of radial velocity (RV) surveys conducted to find exoplanets. In most cases, however, these have not so far been useful for asteroseismology, because these RV data are sparsely and irregularly-sampled. Because we have to pause observations during the day, during poor weather conditions, or simply when targets of higher priority are being observed, we get time series of only a few points and which may have significant and uneven gaps. This introduces a window-function effect: the power spectrum as constructed for example by a Fourier transform, or a Lomb-Scargle Periodogram \citep{lomb,scargle}, is convolved with the Fourier transform of the window function, introducing strong sidelobes adjacent to real frequency peaks and causing crosstalk between adjacent frequency channels. This imposes significant limitations both on the signal-to-noise and frequency resolution of power spectra derived from linear methods such as the Lomb-Scargle periodogram, and in practice makes asteroseismology with these conventional approaches difficult or impossible from the ground for stars with oscillation periods ranging from $\sim 12$~h to $\sim$~a few days.

If we apply nonlinear statistical inference methods this situation can be improved. \citet{brewer2009} show that a system of driven, damped harmonic oscillators such as we encounter in asteroseismology can be statistically modelled as a Gaussian Process, with a covariance kernel consisting of a sum of damped sinusoids. The hyperparameters of this process can be used to retrieve very accurate power spectral density distributions. Unfortunately, this is impractical for long time series, as the computational cost of evaluating the standard Gaussian process likelihood function scales as $\mathcal{O}(n^3)$.

A Gaussian process such as that used by \citet{brewer2009} also
corresponds to the solution, $y(t)$, of the stochastic ODE
\begin{equation}
  \label{eq:GP-definition}
  \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) = \prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t),
\end{equation}
where $p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the
process; $r_i \in \mathbb{C}$, $\Re r_i < 0$ are the roots of the
characteristic equation of the ODE; $b_j \in \mathbb{C}$,
$\Re b_j < 0$ are corresponding roots of the inverse process; and
$\eta(t)$ is a white-noise Gaussian process with
$\left\langle \eta(t) \right\rangle = 0$ and
$\left\langle \eta(t) \eta(t') \right\rangle = \sigma^2 \delta\left(t
  - t' \right)$.  Such a solution is called a CARMA (Continuous
AutoRegressive Moving Average) process.  The constraint that
$\Re r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator
defining the process and its inverse are invertable.  Under the
assumption that the solution, $y(t)$, is real, roots $r_i$ and $b_j$
must either be real or occur in complex conjugate pairs.  The power
spectrum of the process is
\begin{equation}
  \label{eq:CARMA-PSD}
  P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation}
The power spectrum is a rational function of frequency with poles at
$2\pi i f = r_i$ and zeros at $2\pi i f = b_i$; because of the
invertable constraints, these poles and zeros all occur when
$\Im f \neq 0$.  The autocorrelation function corresponding to the
power spectrum in Eq.\ \eqref{eq:CARMA-PSD} is
\begin{equation}
  \label{eq:CARMA-ACF}
  \rho_y (\tau) = \left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle = \sum_{i=1}^p A_i^2 e^{r_i \tau},
\end{equation}
where the RMS amplitudes of the $p$ different modes, $A_i$, are
functions of $\sigma$ and the values of the roots $r_i$ and $b_j$.  If
$q = p-1$, then the $A_i$ are independent, while if $q < p-1$ then
there are interdependencies among the $A_i$.  From either the power
spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the autocorrelation function
in Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a real root $r_i$
represents an exponentially-decaying mode with $r_i$ the rate
constant, while a complex conjugate pair of $r_i$ represent an
oscillatory mode with decay rate $\Re r_i$ and angular frequency
$\left|\Im r_i\right|$.  The former is called an CAR(1) process, an
Ornstein-Uhlenbeck process, or a damped random walk.
\citet{Kelly2014} gives more details on CARMA processes, including
formulae for $A_i$ as a function of $\sigma$, $r_i$, and $b_j$, and
shows how a state-space model of the solution $y(t)$ can be tracked
through a time-series of uncertain observations
$y_k = y\left( t_k \right) + \epsilon_k$, with Gaussian noise
$\epsilon_k$ of zero mean and known (heteroskedastic) variance, using
a Kalman filter to produce a likelihood function
$p\left( y_k \mid \sigma, r_i, b_j \right)$ in
$\mathcal{O}\left( n \right)$ computational time.  Additionally,
\citet{Foreman-Mackey2017} show how a dimensional expansion can be
used to reduce the standard Gaussian process covariance matrix for the
process with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$ to a banded form that can then be
Cholesky-decomposed in the standard Gaussian process likelihood
function in $\mathcal{O}\left( n \right)$ time.

@@

\section{Aldebaran}
Aldebaran is a red giant star with spectral type K5, one of the nearest such stars at a distance of only $19.96 \pm 0.38$~pc \citep{hipparcos}. Its position near the Ecliptic permits the determination of its angular diameter by lunar occultations and by interferometry \citep[$20.58 \pm 0.03$ mas;][]{richichi2005,1979ApJ...228L.111B,brown1979,panek1980}.

>>>>

It is easy to observe Aldebaran and similarly-bright stars with ground-based spectroscopic instruments, typically only requiring short exposures that can be obtained even under adverse observing conditions. There is indeed a considerable archive of such observations already, as a legacy of radial velocity (RV) surveys conducted to find exoplanets. In most cases, however, these have not so far been useful for asteroseismology, because these RV data are sparsely and irregularly sampled. Because we have to pause observations during the day, during poor weather conditions, or simply when targets of higher priority are being observed, we get time series which may have significant and uneven gaps. This introduces a window-function effect: the power spectrum as constructed for example by a Fourier transform, or a Lomb-Scargle Periodogram \citep{lomb,scargle}, is convolved with the Fourier transform of the window function, introducing strong sidelobes adjacent to real frequency peaks and causing crosstalk between adjacent frequency channels. This imposes significant limitations both on the signal-to-noise and frequency resolution of power spectra derived from linear methods such as the Lomb-Scargle periodogram, and in practice makes asteroseismology with these conventional approaches difficult or impossible from the ground for stars with oscillation periods ranging from $\sim 12$~h to $\sim$~a few days.

If we apply nonlinear statistical inference methods this situation can be improved. \citet{brewer2009} show that a system of driven, damped harmonic oscillators such as we encounter in asteroseismology can be statistically modelled as a Gaussian Process, with a covariance kernel consisting of a sum of damped sinusoids. The hyperparameters of this process encode features of the power spectral density distributions and are insensitive to the window function. Unfortunately, this is impractical for long time series, as the computational cost of evaluating the standard Gaussian Process likelihood function scales as $\mathcal{O}(n^3)$.

A Gaussian Process such as that used by \citet{brewer2009} also
corresponds to the solution, $y(t)$, of the stochastic ODE
\begin{equation}
  \label{eq:GP-definition}
  \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) = \prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t),
\end{equation}
where $p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the
process; $r_i \in \mathbb{C}$, $\Re r_i < 0$ are the roots of the
characteristic equation of the ODE; $b_j \in \mathbb{C}$,
$\Re b_j < 0$ are corresponding roots of the inverse process; and
$\eta(t)$ is a white-noise Gaussian Process with
$\left\langle \eta(t) \right\rangle = 0$ and
$\left\langle \eta(t) \eta(t') \right\rangle = \sigma^2 \delta\left(t
  - t' \right)$.  Such a solution is called a CARMA (Continuous
AutoRegressive Moving Average) process.  The constraint that
$\Re r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator
defining the process and its inverse are invertable.  Under the
assumption that the solution, $y(t)$, is real, roots $r_i$ and $b_j$
must either be real or occur in complex conjugate pairs.  The power
spectrum of the process is
\begin{equation}
  \label{eq:CARMA-PSD}
  P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation}
The power spectrum is a rational function of frequency with poles at
$2\pi i f = r_i$ and zeros at $2\pi i f = b_i$; because of the
invertable constraints, these poles and zeros all occur when
$\Im f \neq 0$.  The autocorrelation function corresponding to the
power spectrum in Eq.\ \eqref{eq:CARMA-PSD} is
\begin{equation}
  \label{eq:CARMA-ACF}
  \rho_y (\tau) = \left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle = \sum_{i=1}^p A_i^2 e^{r_i \tau},
\end{equation}
where the RMS amplitudes of the $p$ different modes, $A_i$, are
functions of $\sigma$ and the values of the roots $r_i$ and $b_j$.  If
$q = p-1$, then the $A_i$ are independent, while if $q < p-1$ then
there are interdependencies among the $A_i$.  From either the power
spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the autocorrelation function
in Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a real root $r_i$
represents an exponentially-decaying mode with $r_i$ the rate
constant, while a complex conjugate pair of $r_i$ represent an
oscillatory mode with decay rate $\Re r_i$ and angular frequency
$\left|\Im r_i\right|$.  The former is called an CAR(1) process, an
Ornstein-Uhlenbeck process, or a damped random walk.
\citet{Kelly2014} gives more details on CARMA processes, including
formulae for $A_i$ as a function of $\sigma$, $r_i$, and $b_j$, and
shows how a state-space model of the solution $y(t)$ can be tracked
through a time-series of uncertain observations
$y_k = y\left( t_k \right) + \epsilon_k$, with Gaussian noise
$\epsilon_k$ of zero mean and known (heteroskedastic) variance, using
a Kalman filter to produce a likelihood function
$p\left( y_k \mid \sigma, r_i, b_j \right)$ in
$\mathcal{O}\left( n \right)$ computational time.  Additionally,
\citet{Foreman-Mackey2017} show how a dimensional expansion can be
used to reduce the standard Gaussian Process covariance matrix for the
process with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$ to a banded form that can then be
Cholesky-decomposed in the standard Gaussian Process likelihood
function in $\mathcal{O}\left( n \right)$ time.

@@

\section{Aldebaran}
Aldebaran is a red giant star with spectral type K5, one of the nearest such stars at a distance of only $19.96 \pm 0.38$~pc \citep{hipparcos}. Its position near the Ecliptic permits the determination of its angular diameter by lunar occultations and by interferometry \citep[$20.58 \pm 0.03$ mas;][]{richichi2005,1979ApJ...228L.111B,brown1979,panek1980}. These tight constraints are valuable in breaking degeneracies in stellar modelling and make this an ideal star for asteroseismic characterization (Appendix~/ref{sec:mod}).
